---
title       : Storm data report
author      : ZeitgeberH 

---


Tornadoes is the leading cause of human loss and economic consequence 
========================================================

## Synopsis
Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

## Introduction
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

In this report, we want to address the following questions:

- **Across the United States, which types of events are most harmful with respect to population health?**

- **Across the United States, which types of events have the greatest economic consequences?**

We will use the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database for original data. Data analysis was done in R. Report was written with [Knitr](http://yihui.name/knitr/) in [RStudio](https://www.rstudio.com/).

## Data

The original data comes from [the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database](http://www.ncdc.noaa.gov/data-access).This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. 

[The Reproducible research course](https://www.coursera.org/course/repdata) in Coursera provided a dataset from this database. It comes in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:[Data [47Mb]](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), or you can download it from [my repository](https://github.com/ZeitgeberH/Reproducible-Research).

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

- National Weather Service Storm [Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf ) 
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


## Data Processing

First, we need to figure out which colomns contain the data we want and just load these colomns.So we load the first 2rows of the data from the CSV file.

```{r Figure out columns}
StormData <- read.csv(file="repdata-data-StormData.csv",nrows=2)
head(StormData)
```
We see there are `r ncol(StormData)` columns. What we are interested in this report are following columns:

For event type: **EVETPE**

For population health: **FATALITIES** and **INJURIES**. These are the numbers of fatalities and injuries in a event.

For economic consequences: **PROPDMG** and **CROPDMG**. These are the property damage and crop damage in a event.

Lets get column number of these columns.
```{r}
ColN <- colnames(StormData) ## column names

Ecol<-which(ColN=="EVTYPE")
Fcol<-which(ColN=="FATALITIES") ## col number for fatalitis
Icol<-which(ColN=="INJURIES") ## col number for injuries
Pcol<-which(ColN=="PROPDMG") ## col number for property damage
Ccol<-which(ColN=="CROPDMG") ## col number for crop damage


```

Now, load data
```{r Load data,cache=TRUE}

theList<-rep(list("Character"),length(ColN))

DF<-scan("repdata-data-StormData.csv",what=theList,sep=",",skip=1)

## extract Year of the event
Date<-DF[[2]] 
Date<-as.Date.character(Date,"%m/%d/%Y %H:%M:%S")
Date<-format(Date,"%Y")

## create a dataframe with 6 columns:date, eventype,
## Fatalitis, Injuries, property damage and crop damage
Mydata<-data.frame(Date=Date,Type=DF[[Ecol]],Fd=as.numeric(DF[[Fcol]]),Id=as.numeric(DF[[Icol]]),Pd=as.numeric(DF[[Pcol]]),Cd=as.numeric(DF[[Ccol]]))

## clean it
Mydata<-Mydata[-which(Mydata[,2]=="?"),]
rm(DF)
```
Dimension of this dataset is:
```{r Data_talbe1}
dim(Mydata)
```
Later, when we looked at the data. We found that **THUNDERSTORM WIND** has at least three other names **TSTM WIND** ,**THUNDERSTORMW**, and **THUNDERSTORM WINDS** in this dataset. They should be combined and named as **THUNDERSTORM WIND**
```{r}
Name1<-which(Mydata$Type=="TSTM WIND")
Name2<-which(Mydata$Type=="THUNDERSTORM WINDS")
Name3<-which(Mydata$Type=="THUNDERSTORMW")
Mydata$Type[c(Name1,Name2,Name3)]<-"THUNDERSTORM WIND"
```


We use **dplyr** package to group and summarize data.
```{r dplyr}

suppressPackageStartupMessages(library(dplyr))

Mydata1<-Mydata %.% # group data by Type
         group_by(Type) %.% # summary of total events, total fatalities, total...
         summarise(n=n(),sum_F=sum(Fd),sum_I=sum(Id),sum_P=sum(Pd),sum_C=sum(Cd)) %.%
          mutate(sum_Health=sum_F+sum_I,sum_Damage=sum_P+sum_C,avg_H=sum_Health/n,avg_D=sum_Damage/n)

```


**Mydata1** contains data summary for both human loss and enconomic death across USA between year 1950 and 2005. We arrange the dataframe acorrding to human loss and encomic death respectively. 

```{r }
DF_H<-arrange(Mydata1,desc(sum_Health))
# arrange factor as the order of health record
DF_H$Type <- factor(DF_H$Type, levels=unique(DF_H$Type), ordered=TRUE)


DF_D<-arrange(Mydata1,desc(sum_Damage))
# arrange factor as the order of damage record
DF_D$Type <- factor(DF_D$Type, levels=unique(DF_D$Type), ordered=TRUE)


DF_HA<-arrange(Mydata1,desc(avg_H))
# arrange factor as the order of health record
DF_HA$Type <- factor(DF_HA$Type, levels=unique(DF_HA$Type), ordered=TRUE)


DF_DA<-arrange(Mydata1,desc(avg_D))
# arrange factor as the order of damage record
DF_DA$Type <- factor(DF_DA$Type, levels=unique(DF_DA$Type), ordered=TRUE)


```

Second, we want to know what's the severe events in most recent three years.
```{r Data_table2_top3}        
Date_ordered<-factor(unique(Date))
NYear<-length(Date_ordered)
str(Date_ordered)
Mydata2<-Mydata %.% # group data by Type and Date
         filter(Date==Date_ordered[NYear] | Date==Date_ordered[NYear-1] | Date==Date_ordered[NYear-2]) %.%
         group_by(Date,Type) %.% # summary of total events, total fatalities, total...
         summarise(n=n(),sum_F=sum(Fd),sum_I=sum(Id),sum_P=sum(Pd),sum_C=sum(Cd)) %.% 
         ## Rank by total health loss and economic damage
         mutate(sum_Health=sum_F+sum_I,sum_Damage=sum_P+sum_C,HRank=dense_rank(-sum_Health),DRank=dense_rank(-sum_Damage)) 
          
# extract the higest 3 ranks
Mydata_H<-Mydata2%.%filter(HRank<4)%.%arrange(HRank)
Mydata_H$Type <- factor(Mydata_H$Type, levels=unique(Mydata_H$Type), ordered=TRUE)

Mydata_D<-Mydata2%.%filter(DRank<4)%.%arrange(DRank)
Mydata_D$Type <- factor(Mydata_D$Type, levels=unique(Mydata_D$Type), ordered=TRUE)
```


Besides, we also want to know the trend through 55 years. So we create another dataset which contains the data of leading cause of human loss or economic loss ordered by years.

```{r Data_table3_trend}
Mydata3<-Mydata %.% # group data by Type and Date
         group_by(Type, Date) %.% # summary of total events, total fatalities, total...
         summarise(n=n(),sum_F=sum(Fd),sum_I=sum(Id),sum_P=sum(Pd),sum_C=sum(Cd)) %.%
         mutate(sum_Health=sum_F+sum_I,sum_Damage=sum_P+sum_C) 

DF_HT<-Mydata3%.%  ## arange by total humn loss
        filter(Type==as.character(DF_H[1,1]))%.%
           arrange(Date)   ## select the higest human loss event type

DF_DT<-Mydata3%.%
       filter(Type==as.character(DF_D[1,1]))%.%
       arrange(Date)

```


## Results

### **`r DF_H[1,1]`** is the leading cause of human loss and economic consequences.
First, let's look at first 100 highest human loss and first 100 higest economic loss of event through out USA across 1950 and 2005. I made a table using **googleVis** package. 

```{r Table1, results='asis'}

n<-100
## It seems google  table doesn't work with Factors.
c<-data.frame(Event=as.character(DF_H[1:n,1]),N=DF_H[1:n,2],Total_Human_Loss=DF_H[1:n,7],Average_Human_Loss=DF_H[1:n,9],Total_Economic_Loss=DF_H[1:n,8],Average_Economic_Loss=DF_H[1:n,10])

suppressPackageStartupMessages(library(googleVis))

op <- options(gvis.plot.tag='chart')

Table <- gvisTable(c, options=list(width = 800, height = 300,page='enable'))

print(Table,"chart")
```
#### Table 1: Total human loss and economic loss across USA between year 1950 and 2005

Looking at data in Table1, It is obvious that **`r DF_H[1,1]`** is the leading cause of human loss and economic loss.

To domontrate this, let's make a simple bar plot using gglot2.
```{r Multiplot }
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  suppressPackageStartupMessages(library(grid))

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

```{r Figure1, fig.width=12, fig.align='left'}
suppressPackageStartupMessages(library(ggplot2))
p1<-ggplot(DF_H[1:5,],aes(Type,sum_Health))+geom_bar(stat="identity",color="black",fill="red")+xlab(label="Event Type")+ylab(label="Total human loss")+ ggtitle('A') +theme(plot.title=element_text(size=14,hjust=0),axis.text.x = element_text(angle = 30, hjust = 1))#plot first 5 highest health records

p2<-ggplot(DF_HA[1:5,],aes(Type,avg_H))+geom_bar(stat="identity",color="black",fill="red")+xlab(label="Event Type")+ylab(label="Average human loss per event")+ ggtitle('B') +theme(plot.title=element_text(size=14,hjust=0),axis.text.x = element_text(angle = 30, hjust = 1))#plot first 5 highest health records

p3<-ggplot(DF_D[1:5,],aes(Type,sum_Damage))+geom_bar(stat="identity",color="black",fill="blue")+xlab(label="Event Type")+ylab(label="Total economic loss")+ ggtitle('C') +theme(plot.title=element_text(size=14,hjust=0),axis.text.x = element_text(angle = 30, hjust = 1))#plot first 5 highest damage records

p4<-ggplot(DF_DA[1:5,],aes(Type,avg_D))+geom_bar(stat="identity",color="black",fill="blue")+xlab(label="Event Type")+ylab(label="Average economic loss per event")+ ggtitle('D') +theme(plot.title=element_text(size=14,hjust=0),axis.text.x = element_text(angle = 30, hjust = 1))#plot first 5 highest damage records

multiplot(p1,p2,p3, p4,cols=2)
```
#### Figure 1. Human loss and economic loss across USA between year 1950 and 2005

Again, we see that **`r DF_H[1,1]`** is the leading cause of human loss and economic loss.


### Since 2009, TORNADO is the leading cause of human loss, while THUNDERSTORM WIND had highest economic consequence.

```{r Table2_Top3, results='asis'}
c<-data.frame(Date= Mydata_H[,1],Event=as.character(Mydata_H[,2]),N= Mydata_H[,3],Total_Human_Loss= Mydata_H[,8],HRank=Mydata_H[,10])#,DRank=Mydata2[,11])
c2<-data.frame(Date= Mydata_D[,1],Event=as.character(Mydata_D[,2]),N= Mydata_D[,3],Total_Economic_Loss= Mydata_D[,9],DRank=Mydata_D[,11])#,DRank=Mydata2[,11])

Table1 <- gvisTable(c, options=list(width = 650, height = 300,page='enable'))
Table2 <- gvisTable(c2, options=list(width = 650, height = 300,page='enable'))
Table12<-gvisMerge(Table1,Table2,horizontal=FALSE)
print(Table12,"chart")
```


```{r Figure2, fig.width=18, fig.align='left'}
p1<-ggplot(Mydata_H,aes(Date,sum_Health,fill=Type))+geom_bar(stat="identity",color="black",position=position_dodge())+xlab(label="Year")+ylab(label="Total human loss" )+ggtitle("Human loss for each year") 

p2<-ggplot(Mydata_D,aes(Date,sum_Damage,fill=Type))+geom_bar(stat="identity",color="black",position=position_dodge())+xlab(label="Year")+ylab(label="Total Economic loss" )+ggtitle("Economic loss for each year") 

multiplot(p1,p2,cols=2)
```
Figure2. Top 3 of events during 3 most recent years.


### Annual number of **`r DF_H[1,1]`** seems increase since 1950.

Now, we look at the trend of **`r DF_H[1,1]`** events.

```{r Table3, results='asis'}

Table <- gvisTable(DF_HT, options=list(width = 600, height = 280,page='enable'))

print(Table,"chart")
```
#### Table 3: Summary of **`r DF_H[1,1]`** for  human loss and economic loss across USA since 1950.

If we arrange  the above talbe as the total fatalities (column sum_Health) with descending order (left click the column). We find year 1974, 2011 and 1953 had the highest number of human loss. These were the 1974,2011 and 1953 [super outbreaks](http://en.wikipedia.org/wiki/Super_Outbreak).

Let's plot the Annual total number of **`r DF_H[1,1]`** reported since 1950.

```{r Figure3,, fig.width=12, fig.align='left'}
ggplot(DF_HT,aes(Date,n))+geom_bar(stat="identity",color="black",fill="red")+xlab(label="Year")+ylab(label="Annual Total Number" )+theme(axis.text.x = element_text(angle = 45, hjust = 1))#plot first 5 highest health records

```
Figure 3. The annual total number of  **`r DF_H[1,1]`** reported in the U.S. since 1950

It seems annual total number of **`r DF_H[1,1]`** increases over the years. However, we must admit that the data only reflects the reported case, which should be related the actual number of occurance but biased by how the data was reported. See [here](http://www.wunderground.com/climate/extreme.asp?MR=1) for more explanations. 

## Notes
You may have other sections in your analysis, but Data Processing and Results are required.

The analysis document must have **at least one figure** containing a plot.

Your analysis must have **no more than three figures**. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default setting in knitr)


```{r clean work}
## Set options back to original options
options(op)
```

## Resources

